A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-07-10 text/manuscript.Rmd
split text into smaller files for ease of editing

## Abstract
Cancer is a disease of cellular metabolism and other basic biochemical processes
common to all cells. However, molecular signatures remain largely
indistinguishable from tissue-of-origin, and general, unifying principles remain
obscure. To address this challenge, we reduced mutation data to disruptions in a
select set of pathways pointing to basic cell biology, from DNA replication to
cellular communication. Using dimension reduction techniques, we assigned tumor
samples into twelve classes distinct from their tissues-of-origin and largely
free of bias from mutational burden or clinical stage. We showed that these
tumor classes vary in prognosis by modeling relative risk of death in different
classes and cancer types. We identified class-specific mutations in different
tissues, demonstrating how tissue-specific signatures contribute to common
cellular phenotypes. Moreover, germline risk mutations involved in genome
stability, such as BRCA1 and Rb1, are equally distributed among our classes,
somewhat contrary to the expectation that such genes are hallmarks of molecular
subtypes. We also investigated pathways that distinguish between metastatic and
non-metastatic tumors, and found that most meaningful differences are
class-specific. Most metastatic enriched pathways from one class in metastatic
tumors are class-specific pathways from another class, suggesting phenotypic
convergence, and potentially highlighting key pathways to focus on
therapeutically for prevention of class-specific relapse. In summary, our
observations suggest that some genes act as drivers of genome instability in all
classes, while different sets of class-specific genes confer increased fitness
in a combinatorial fashion, eventually converging on one or more end-stage
disease phenotypes. Thus, we present a coherent picture of global tumor biology,
and explain how common cellular dysfunction might arise from tissue-specific
mutations.

@@

Recent advances in technology have made the study of cancer possible at
unprecedented levels of detail and scale. Array and next-generation sequencing
technologies have made it possible to assay germline and somatic mutations,
expression profiles, DNA methylation, and copy number variations. To take
advantage of these tools several large consortia, including the cancer genome
atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG), sequenced large
numbers of tumors and collected data from multiple assays that could be analyzed
together, with the goal of integration and increased understanding of the
mechanisms of cancer. These efforts have focused on recognizing and quantifying
the importance of genes in cancer etiology as the result of selection processes,
or classification of cancers relative to common patterns of aberration in
multiple assays.

Considerable progress has been made analyzing these data. Statistical analyses
identify hundreds of global and tissue-specific cancer driver genes
[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
when genes are mutated at a greater rate than expected due to chance. Based in
part on studies like these, it has been estimated that on the order of three
mutations in key oncogenes and/or tumor suppressors would be sufficient to
transform a normal cell to a cancerous state [@vogelsteinPathCancerThree2015].

Other studies aim at providing a comprehensive view of mutations, gene
expression and genomic signatures, with the express goal of understanding the
common themes of all cancer independent of tissue of origin. Such a view of
cancer as a malady affecting common cellular processes has long been a goal of
the field and is perhaps best characterized in essays by @weinberg2000; @hallmarks_II (see also
Discussion). The first of these studies, (the
“Pan-Cancer-12” analysis) identified 11 subtypes from 12 cancer types,
using integrative analysis with co-equal weighting of gene expression,
methylation, copy number and proteomics data
[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
tissue-of-origin is the predominant driving factor, though a significant
proportion of tumors ($10\%$) could be reclassified independent of
tissue-of-origin. In a second study involving 33 cancer types and a much greater
number of tumors using a newer algorithm (iCluster)
[@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that
could be further subdivided into organ specific groups, including
pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/
squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].

More recently, attempts have been made to consider pan-cancer mutations and
other data outside the context of enrichment at the gene level. Horn _et al._
[@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the
sparseness of mutations, allowing the potential importance of a gene to be
influenced by the aggregate of positive selection in its nearest network
neighbors. This method is powerful especially when applied to a small number (on
the order of hundreds) of tumors. However, the larger problem of cancer etiology
remains: what mutations are required, in what order, and are there specific
combinations of delinquent cellular processes sufficient to cause cancer? If
there is a common logic to cancer, how is it to be detected underneath the
overwhelming cell-of-origin signal?

In this study, we attempt to address these questions by focusing on discrete
molecular pathways as the unit of disruption in cancer. Our approach strips all
cell-type-specific information from the mutation data and generalizing
gene-level information to non-tissue specific cell-biological pathway
phenotypes. We use this rubric to evaluate all cancers and show that,
surprisingly, tumors may have distinct phenotypes completely independent of
their tissues of origin.

@@

### Classification of tumors independent of tissue-of-origin

The basic problem in tumor research of this kind, namely that the search space,
~20,000 protein-coding genes is greater than the number of tumor samples
[@hornNetSigNetworkbasedDiscovery2018]. And that is not accounting for the fact
that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
elements exist with oncogenic or tumor suppressor potential. Hence, we sought to
investigate cancer as a disease of basic cellular and biochemical pathways,
which we accomplished by translating gene-level mutations into pathway level
disruptions. Importantly, we selectively removed pathways that lacked generality
as cellular processes or that could be specifically assigned to differentiated
cell types, while excluding others.

We hypothesize that cancer is the result of dysfunction from a limited number of
basic cellular processes that are common to all living human cells, and thus all
tumors belong on a spectrum of dysfunction on this axis regardless of tissue of
origin. An alternative hypothesis is that every tumor belongs to one of a large
number of syndromes which are unique to each tissue of origin, that share some
mechanisms and superficial similarity in treatment strategies. The latter idea
has gained momentum in recent years since the publication of Encode, Roadmap and
TCGA consortium papers, perhaps owing to the intractable complexity of genomic
data in all its forms. The inescapable conclusion thus far is that all these
data point to tissue-of-origin as the most important driver of tumor behavior.
Our observations contrast with this view, and instead support an interpretation
in which all tumors manifest one of a limited number of metastable phenotypes
represented by our classes (**figure 1a-k**), with increased fitness regardless
of tissue of origin, which then converge on end stage cancer by accumulating
complementary mutations to the other classes (**figure 1l** and **Table 2**).

We find that most tumors segregate into one of 12 biologically distinct
clusters, regardless of the cancer type (**figure 1h** and **supplementary
figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
arm clusters 9 and 10, and breast cancers being excluded from the same cluster,
see **supplementary figure S1**) are unevenly distributed toward a specific
class, we could not identify cancer type that was exclusive to a single tumor
class. Somewhat surprisingly to us, this finding extends to histological
subtypes of breast, head and neck cancers, leukemias, _etc._ (see **Figure 1i**
and **Supplemental Figure S2**). Indeed, it would be surprising if we did not
observe bias for some cancers, as it has long been known that certain treatment
regimens are likely to work for a majority of patients of a given cancer type.
Nonetheless, the basis for some tumors being treatment-refractory in spite of
clinical indications remains elusive. Our analysis suggests the possibility of
more general mechanisms underlying tumor evolution regardless of tissue,
histological or cellular subtype and could form a basis for future separation of
patients into different precision treatment regimens.

It could be that the classes we identified are consistent with a signature of
overall disease progression, such that successive, adjacent classes on the UMAP
projection exhibit increasing specificity. We identified no such trend (see
**figure 1k**) and each class instead was associated with unique combinations of
pathways (as in **figure 3**). In our view, these observations suggest that
cancers have available to them a limited number of stable configurations with
respect to pathway disruption which are consistent with increased fitness and
therefore positive selection.

Finally, we must remark on two of the classes, 1 and 8, which for which we did
not find distinctive associations with pathways. Class 1 has a relatively low
level of mutated pathways overall, although it is broadly enriched in many of
the same tumor-promoting pathways common to the other groups. This likely does
not reflect a true difference in stages of tumor evolution, as our data show
clearly that this class is as likely to contain stage IV as it is to contain
stage I tumors. We speculate this class represents a group of tumors with large
aberrations in methylation, copy number or other genome rearrangements which we
excluded from the present analysis. In the future, we hope to incorporate these
other data into a pathway-centric analysis as we have done here for point
mutations and indels. Class 8 represents a group of hyper-mutated tumors that
harbor so many mutations that virtually no pathway is unaffected. This group
highlights one of the limitations of the approach that we’ve taken, in that it
is likely that a significant fraction of the samples flagged as “mutant” for
each pathway are actually unaffected, consistent with those samples have low
information content. Similar to class 1 we speculate that a future analysis of
other data types, including improved filtering of mutations, may resolve these
tumors into more meaningful categories.

### Incompleteness of the pathway disruption data

The fact that we are likely excluding many defects (under chromosomal
abberations, amplifications, demethylation and so on) also raises the
possibility that our classes are incomplete or may change at a future time when
we can account for them. For now at least, we think that the point mutation data
are representative of the most important pathways.

### Estimates of survival reveal pathway-dependent differences

By modeling CDC longevity data as a baseline risk function we showed that each
of our pathway disruption classes exhibit cancer-type specific effects on
expected survival (**figure 5**). Contrary to what one might expect at the
outset, we found that models assuming different survival based on class _and_
cancer type fit better than models in which these variables are independent
factors. However, considering that within each cancer type there are different
clinical standards of care, and even within classes of drugs the preferred
treatment can very between cancers, it makes sense in hindsight that they should
all be different. Contrast the situation with ovarian vs breast cancer, which
are both hormonally driven cancers, for example. Ovarian cancer has but one main
treatment, platinum, whereas breast cancer patients have a variety of treatment
regimens based on molecular subtype and other factors. It is easy to see how
standard of care for one cancer and class could produce different outcomes than
the same class of a different cancer type under the circumstances.
Unfortunately, given the diversity drug classes and treatments (including
non-chemical treatments such as surgery and palliative care), we lack sufficient
power in the current study to explore these variables in the GDC data. It is our
hope that more focused future studies will help to distinguish between
treatment-specific effects or survival given different pathway disruption
classes.

### Implications for the evolution of cancer

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, class-specific
mutations and metastasis. General cancer promoting processes must include
factors that relate to genome stability either directly or indirectly. Such
pathways are disrupted in all or most of the defined classes (see **figures 2**
and **supplementary figure S1**) and often are the result of abberations
involving common driver genes such as TP53, BRCA1/2, MMR genes, mitotic
checkpoints, Cohesion complexes, _etc._ (**supplementary figure S3** and
**supplemental file 1**). Class-specific evolution must involve the acquisition
of combinatorial disruptions to pathways that may individually be harmful but
together produce more specialized cancer phenotypes and increased fitness (as in
**figure 4** and **tables 1** and **ST2**). Importantly, our work does not
distinguish between the order in which these mutations should accumulate.
However, since many of the genes in the non-specific pathways involve genome
stability, it is reasonable to hypothesize that such mutations promote the
acquisition of class specific defects, and natural selection produces the
classes we observed.

### On metastasis as a convergence of phenotypes

The class-specific metastatic pathway enrichment we observe (in **table 2**) may
be a result of the pathways directly promoting metastatic phenotypes, or
alternatively that such mutations are more likely to turn metastatic because
they acquire a stronger or more complete oncogenic phenotype, since (again) we
don’t know the order in which they were acquired. The fact that most metastatic
enrichment is cluster-specific and has a tendency to overlap with
cluster-specific pathways from non-metastatic tumors of neighboring clusters
favors the latter explanation, and suggests a mechanism of complementarity in
which newly acquired mutations converge on a single deadly phenotype that
captures most or all of the critical features of end-stage cancer. If this is
the case, we speculate that once we discover the vulnerabilities of each cluster
we might also reasonably predict which pathways are susceptible to further
selection and therefore target them  therapeutically at the same time to prevent
further adaptation.

>>>>

## Abstract

```{r child='abstract.Rmd'}
```

@@

```{r child='introduction.Rmd'}
```

@@

```{r child='discussion.Rmd'}
```
